---
title: "John Hopkins COVID Data Analysis"
author: "Eric"
date: "2024-04-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

```{r JHU Data Extract, include = TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
filenames <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,filenames)
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```

```{r Tidy and clean global data, include=TRUE}

global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>%  select(-c(Lat, Long))
global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>%  select(-c(Lat, Long))
global <- global_cases %>% full_join(global_deaths) %>% rename(Province_State = 'Province/State', Country_Region = 'Country/Region') %>% mutate(date = mdy(date))
global <- global %>% filter(cases > 0)
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)
uid_lookup <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup)
global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key.x)
```

```{r Tidy and clean US data, include = TRUE}
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key),names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population),names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% select(-c(Lat, Long_)) %>% mutate(date = mdy(date))
US <- US_cases %>% full_join(US_deaths)
```

```{r create additional US data, include=TRUE}
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
US_totals <- US_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
```

## COVID Data Visualizations
The following visualizations represent analysis of the data extracted from John Hopkins recorded COVID data.
```{r visualize data, include=TRUE}
US_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = "COVID 19 in the US", y = NULL)
state <- "Oregon"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("COVID 19 in ", state), y = NULL)
```

## Bias Identification
The data used in this analysis was data provided by John Hopkins and identified COVID 19 cases and deaths which were confirmed to be associated to COVID 19. There is likely data missing from this analysis where citizens did not report or did not go through traditional health system mechanisms which would have ensured there cases were recorded in this data.

```{r model data, include=TRUE}
US_by_state_sum <- US_by_state %>% group_by(Province_State) %>% summarise(cases = sum(cases), deaths = sum(deaths), Population = max(Population))
US_by_state_sum <- US_by_state_sum %>% mutate(cases_per_thou = cases *1000 / Population, deaths_per_thou = deaths *1000 / Population)
US_by_state_sum <- US_by_state_sum %>% filter(Population != 0)
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_by_state_sum)
US_by_state_sum_w_pred <- US_by_state_sum %>% mutate(pred = predict(mod))
US_by_state_sum_w_pred %>% ggplot() + geom_point(aes(x=cases_per_thou, y = deaths_per_thou), color = 'blue')+ geom_point(aes(x = cases_per_thou, y = pred), color = 'red')+theme(legend.position = "bottom", axis.text.x = element_text(angle = 90)) + labs(title = 'COVID 19 Deaths by Cases Model', y = 'Deaths per thousand', x = 'Cases per thousand')
```

```{r}
sessionInfo()
```